---
title: "Chapter 3: Sampling the Imaginary"
output: 
  html_document: 
    highlight: kate
    theme: journal
---

```{r}
library(ggplot2)
library(dplyr)
library(rethinking)
```

# Easy

```{r}
p_grid <- seq(from = 0,
              to = 1,
              length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6 , size = 9 , prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample(p_grid,
                  prob = posterior,
                  size = 1e4,
                  replace = TRUE)
```

3E1. How much posterior probability lies below p = 0.2?

```{r}
sum(samples < 0.2)/length(samples)
```

3E2. How much posterior probability lies above p = 0.8?

```{r}
sum(samples > .8)/length(samples)
```

3E3. How much posterior probability lies between p = 0.2 and p = 0.8? 

```{r}
sum(samples > 0.2 & samples < 0.8)/length(samples)
```

3E4. 20% of the posterior probability lies below which value of p?

```{r}
quantile(samples, 0.2)
```

3E5. 20% of the posterior probability lies above which value of p?

```{r}
quantile(samples, 0.8)
```
3E6. Which values of p contain the narrowest interval equal to 66% of the posterior probability?

```{r}
HPDI(samples, prob = 0.66)
```

3E7. Which values of p contain 66% of the posterior probability, assuming equal posterior probabil- ity both below and above the interval?

```{r}
PI(samples, prob = 0.66)
```

# Medium

3M1. Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

```{r}
p_grid <- seq(from = 0,
              to = 1,
              length.out = 1000)
prior <- rep(1, 1000)
likelihood <- dbinom(8, size = 15, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```

3M2. Draw 10,000 samples from the grid approximation from above. Then use the samples to cal-
culate the 90% HPDI for p.

```{r}
set.seed(100)
samples <- sample(p_grid,
                  prob = posterior,
                  size = 1e4,
                  replace = TRUE)
HPDI(samples, .9)
```

3M3. Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?

```{r}
PPD <- rbinom(n = length(samples), size = 15, prob = samples)
mean(PPD == 8)
```

3M4. Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses.

```{r}
PPD <- rbinom(n = length(samples), size = 9, prob = samples)
mean(PPD == 6)
```

3M5. Start over at 3M1, but now use a prior that is zero below p = 0.5 and a constant above p = 0.5. This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value p = 0.7.

Q1.

```{r}
p_grid <- seq(from = 0,
              to = 1,
              length.out = 1000)
prior <- as.numeric(!p_grid < 0.5)
likelihood <- dbinom(8, size = 15, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```

Q2. 

```{r}
set.seed(100)
samples <- sample(p_grid,
                  prob = posterior,
                  size = 1e4,
                  replace = TRUE)
HPDI(samples, .9)
```

Q3.

```{r}
PPD <- rbinom(n = length(samples), size = 15, prob = samples)
mean(PPD == 8)
```

Q4.

```{r}
PPD <- rbinom(n = length(samples), size = 9, prob = samples)
mean(PPD == 6)
```

# Hard

```{r}
data(homeworkch3)
```

3H1.

```{r}
boys <- sum(birth1, birth2)
births <- length(birth1) + length(birth2)
girls <- births - boys
prior <- rep(x = 1, times = length(p_grid))
likelihood <- dbinom(x = boys, size = births, prob = p_grid)
posterior <- prior * likelihood
posterior <- posterior/sum(posterior)
plot(posterior ~ p_grid, type = "l")
p_grid[which.max(posterior)]
```

3H2.

```{r}
samples <- sample(x = p_grid, size = 1e4, replace = TRUE, prob = posterior)
HPDI(samples, c(.5, .89, .97))
```


3H3.

```{r}
PPD <- rbinom(n = 1e4, size = births, prob = samples)
dens(PPD)
abline(v = boys)
```

3H4.

Two ways to interpret. Can draw from the samples we have already.
```{r}
PPD <- rbinom(n = 1e4, size = 100, prob = samples)
dens(PPD)
abline(v = sum(birth1))
```

Or set up a new likelihood, posterior and PPD for just first births.

```{r}
likelihood <- dbinom(x = sum(birth1), size = 100, prob = p_grid)
posterior <- likelihood * prior
posterior <- posterior/sum(posterior)
samples <- sample(p_grid, size = 1e4, replace = TRUE, prob = posterior)
PPD <- rbinom(n = 1e4, size = 100, prob = samples)
dens(PPD)
abline(v = sum(birth1))
```

3H5.

```{r}
girl_first <- !birth1
has_big_sister <- birth2[girl_first]
PPD <- rbinom(n = 1e4, size = sum(girl_first), prob = samples)
dens(PPD)
abline(v = sum(has_big_sister))
```

